---
title: "Model fits to experimental data"
output:
  pdf_document: default
  html_notebook: default
---

# Load required packages # 

```{r message=TRUE, warning=TRUE, include=FALSE}
library(rjags)
library(dclone)
library(tidyverse)
library(readxl)
library(MCMCvis)
library(emdbook) # rzinbinom()
```

# Load the data # 

Load the data from the excel sheet. Truncate the data at `obs_day >= 50` to include only observations from the endemic phase. 

```{r warning=FALSE, messages = FALSE}
endemic_df <- read_excel(
    "/Users/leilakrichel/Desktop/Manuscript/ExperimentalData/ExperimentalData_2022.xlsx") %>% 
  filter(obs_day >= 50) %>% 
  select(1:5)

population_ids <- endemic_df$population %>% unique()
```

# Logistic regression model #

This model is fit to binary infection status data (individual is infected or uninfected) to estimate endemic infection prevalence. 

## Prepare the data for jags ##

Convert the data into a list with the following categories for each observation:

- treatment: 2 levels (constant or variable)
- population: 12 levels (replicate population)
- Ntreatment: number of treatments
- Npopulation: number of populations
- y: 2 levels (infection status, 1 is infected, 0 is uninfected)

```{r}
endemic_jags <- 
  list(
    "treatment" = ifelse(endemic_df$population %in% population_ids[1:6], 1, 2), 
    "population" = as.numeric(as.factor((endemic_df$population))), 
    "Ntreatment" = 2,
    "Npopulation" = 12, 
    "y" = ifelse(endemic_df$n_clusters > 0, 1, 0)
    )
```

## Define the logistic regression model ##

```{r}
logit_model <- function() {
  
  for(i in 1:length(y)) { 
    
    y[i] ~ dbern(p[i])
    logit(p[i]) <- bT[treatment[i]] + aP[population[i]]
    
  }
  
  #Priors
   
  for (j in 1:Ntreatment) {
    
       bT[j] ~ dnorm(0, 0.5) # treatment effect 
    
     }
  
  for(k in 1:Npopulation) { 
 
    aP[k] ~ dnorm(0, tauP) # population effect
    
  }
  
  tauP <- 1/(sigP * sigP)
  sigP ~ dunif(0, 100)
  
  
}
```

## Run the logistic regression model ##

```{r eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
logit_run <- jags.fit(
  data = endemic_jags,
  params = c("bT", "aP"),
  model = logit_model,
  n.chains = 4)
```

## Save the logistic regression model ##

```{r, eval = FALSE}
save(logit_run, 
     file = "/Users/leilakrichel/Chp1/RData_Manuscript/LogitExp_Fig2.RData")
```

# Zero-inflated negative binomial model #

This model is fit to infection burden data to estimate the distribution of infection burden at the endemic phase. 

## Prepare the data for jags ##

Convert the data into a list with the following categories for each observation:

- treatment: 2 levels (1 is constant, 2 is variable)
- population: 12 levels (1:6 are constant replicate populations, 7:12 are variable replicate populations)
- Ntreatment: number of treatments
- Npopulation: number of populations
- y: observations of infection burden 

```{r}
burden_jags <-
  list(
    "treatment" = ifelse(endemic_df$population %in% population_ids[1:6], 1, 2), 
    "Ntreatment" = 2,
    "Npopulation" = 12,
    "population" = as.numeric(as.factor((endemic_df$population))), 
    "y" = endemic_df$n_clusters) 
```

## Define the zero-inflated negative binomial model ##

Three parameters are being estimated from infection burden data:

(1) **m, mean of negative binomial process:** varies by treatment, bT, and population, aP
(2) k, overdisperesion parameter: varies by treatment
(3) theta, probability of zero-inflation: varies by treatment

```{r}
ZINB_model <- function() {
  
  for(i in 1:length(y)) { 
    
    z[i] ~ dbern(theta[treatment[i]]) # ZI likelihood function 
    y[i] ~ dnegbin(p[i], k[treatment[i]]) # NB likelihood function
    
    p[i] <- k[treatment[i]]/(k[treatment[i]] + lambda.zinb[i])
    lambda.zinb[i] <- z[i] * m[i] 
    
    m[i] <- exp(bT[treatment[i]] + aP[population[i]]) 

  }
  
  # Priors
  
  for (j in 1:Ntreatment) { 
    
    bT[j] ~ dnorm(3, 0.5) # treatment effect 
    k[j] ~ dgamma(0.01, 0.01) 
    theta[j] ~ dunif(0.000001, 0.9999999)
    
  }
  
  for(l in 1:Npopulation) { # for each population
    
    aP[l] ~ dnorm(0, tauP) # population effect 
    
  }
  
  tauP <- 1/(sigP * sigP) # precision 
  sigP ~ dunif(0, 100)
}
```

## Run the ZINB model ##

```{r message=FALSE, warning=FALSE}
ZINB_run <- jags.fit(
  data = burden_jags,
  params = c("bT", "aP", "k", "theta"),
  n.chains = 4,
  model = ZINB_model)
```

## Treatment-level posterior predictions of ZINB ##

Use the fitted model `ZINB_run` to get distributions of mean infection burden and mean endemic prevalence in constant and variable temperature treatments. These distributions will be derived by "simulating" the experiment. 

```{r}
(zinb_posteriors <- 
  as.data.frame(MCMCchains(ZINB_run)) %>% 
  mutate(
    chain = c(rep(1, 5000), 
              rep(2, 5000), 
              rep(3, 5000), 
              rep(4, 5000)),
    observation = rep(1:5000, 4)
    ) %>% 
   mutate(across(1:14, exp)) %>% 
  gather(key = "parameter", value = "value", 1:18))

replicate_population <- as.data.frame(MCMCchains(ZINB_run)) %>% 
  select(1:12) %>% 
  names()

treatment <- as.data.frame(MCMCchains(ZINB_run)) %>% 
  select(13:14) %>% 
  names()

overdispersion <- as.data.frame(MCMCchains(ZINB_run)) %>% 
  select(15:16) %>% 
  names()

zero_inflation <- as.data.frame(MCMCchains(ZINB_run)) %>% 
  select(17:18) %>% 
  names()
```

Initialize storage objects: arrays will contain a matrix for constant and variable treatments and the matrix will contain the generated time-series of endemic prevalence and infection burden.

```{r}
zinb_prevalence <- replicate(2, vector(mode = "numeric", length = 5000), simplify = FALSE)
zinb_burden <- replicate(2, vector(mode = "numeric", length = 5000), simplify = FALSE)
zinb_fitted <- replicate(2, matrix(NA, nrow = 5000, ncol = 100), simplify = FALSE)
```

To simulate the experiment using the ZINB model, the following steps are coded in the for loop:

1. Randomly choose a chain in the constant or variable treatment.
2. Sample a random observation in the chosen chain to get a set of parameters (mu, k, theta). 
4. Use the set of parameters to draw a sample of 12 "individuals" from the fitted ZINB distribution. 
5. Calculate prevalence in the sample and mean infection burden. 

This will compute distributions of endemic prevalence and mean infection burden in constant and variable treatments. 

```{r eval = FALSE}
for (i in 1:length(zinb_prevalence[[1]])) { 
  for (j in 1:2) { # for each treatment
    for (m in 1:100) {

      set.seed(i * j * m)
      # choose set estimated parameters from ZINB model fit
      ## constant
      
      zinb_params <- zinb_posteriors %>%
        filter(chain == sample(1:4, 1, replace = FALSE)) %>%
        filter(observation == sample(1:5000, 1, replace = FALSE)) %>%
        filter(parameter %in% c(
                                treatment[j],
                                overdispersion[j],
                                zero_inflation[j])) %>%
        select(value) %>%
        unlist() 
      
      # sample from the fitted ZINB distribution
      ## constant
      samples <- rzinbinom(
        150,
        zprob = 1 - zinb_params[3],
        size = zinb_params[2],
        mu = zinb_params[1]
      )
      
      samples2 <- rzinbinom(
        5000,
        zprob = 1 - zinb_params[3],
        size = zinb_params[2],
        mu = zinb_params[1]
      )

      # compute endemic prevalence and mean infection burden
      zinb_prevalence[[j]][i] <- (samples[samples != 0] %>% 
                                  length()) / 150
      
      zinb_burden[[j]][i] <- samples %>% mean()
      zinb_fitted[[j]][, m] <- samples2


      
}}}
```

## ZINB predictions versus experimental observations of endemic prevalence ##

Use the fitted model `ZINB_run` to get distributions of mean infection burden and mean endemic prevalence for each replicate population in constant and variable temperature treatments. These distributions will be derived by "simulating" the experiment (i.e., 26 time steps, 6 replicates per treatment). 

Initialize storage objects. 
```{r}
prevalence_zinbpreds <- matrix(NA, nrow = 26, ncol = 12)
burden_zinbpreds <- matrix(NA, nrow = 26, ncol = 12)
```

Calculate the number of individuals to sample from each replicate population in each treatment at each time step.

```{r}
nt_samples <- matrix(NA,
                     nrow = 26,
                     ncol = length(populations_ids))

for (i in 1:length(populations_ids)) { 

  nt_samples[, i] <- endemic_df %>%
    filter(obs_day != 64) %>% # day 192 of experiment
    filter(population == populations_ids[i]) %>%
    group_by(obs_day) %>%
    summarize(sample_size = n()) %>%
    select(sample_size) %>%
    unlist() 
 } 
``` 

Sample from the fitted ZINB model using the following steps coded in the for loop:

1. Randomly choose a chain in the constant or variable treatment.
2. Sample a random observation in the chosen chain to get a set of parameters (mu, k, theta) for a given replicate population. 
4. Use the set of parameters to draw a sample of individuals from the fitted ZINB distribution. 
5. From the sample calculate endemic prevalence and mean infection burden. 

```{r, eval = FALSE}
for (i in 1:length(replicate_population)) {
    for (m in 1:26) {

      set.seed(i * m)
      # for rep population i: 
      # choose set of estimated parameters from ZINB model fit
      
      if (i <=6) { # constant populations
      zinb_params <- zinb_posteriors %>%
        filter(chain == sample(1:4, 1, replace = FALSE)) %>%
        filter(observation == sample(1:5000, 1, replace = FALSE)) %>%
        filter(
          parameter %in% c(replicate_population[i],
                           treatment[1],
                           overdispersion[1],
                           zero_inflation[1])) %>%
        select(value) %>%
        unlist() 
      } 
    
      else { # variable populations
          
           zinb_params <- zinb_posteriors %>%
        filter(chain == sample(1:4, 1, replace = FALSE)) %>%
        filter(observation == sample(1:5000, 1, replace = FALSE)) %>%
        filter(
          parameter %in% c(replicate_population[i],
                           treatment[2],
                           overdispersion[2],
                           zero_inflation[2])) %>%
        select(value) %>%
        unlist() 
           
           }
      
      # sample from the fitted ZINB distribution
      samples <- rzinbinom(
        nt_samples[m, i],
        zprob = 1 - zinb_params[4],
        size = zinb_params[3],
        mu = zinb_params[2] + zinb_params[1]
      )
      
      # compute endemic prevalence and mean infection burden
      prevalence_zinbpreds[m, i] <- (samples[samples != 0] %>% 
                                  length()) / nt_samples[m, i]
      
      burden_zinbpreds[m, i] <- samples %>% mean()
      
}}
```

## Save the ZINB model for figures 4 and 5 ##

```{r eval = FALSE}
save(ZINB_run, # zinb model fit 
     zinb_posteriors, zinb_prevalence, zinb_burden, zinb_fitted, # treatment-level
     prevalence_zinbpreds, burden_zinbpreds, # replicate-level
     file = "/Users/leilakrichel/Chp1/RData_Manuscript/ZINBExp_Fig4.RData")
```

# Negative binomial model #

This model is fit to infection burden data to estimate the distribution of infection burden at the endemic phase. 

## Prepare the data for jags ##

Use the same data from the ZINB model. 

```{r}
burden_jags %>% 
  glimpse()
```

## Define the negative binomial model ##

Two parameters are being estimated from infection burden data:

(1) **m, mean of negative binomial process:** varies by treatment, bT, and population, aP
(2) **k, overdisperesion parameter:** varies by treatment

```{r}
NB_model <- function() {
  
  for(i in 1:length(y)) { # for each observation [i]
    
    y[i] ~ dnegbin(p[i], k[treatment[i]]) 
    
    p[i] <- k[treatment[i]] / (k[treatment[i]] + lambda[i])
  
    lambda[i] <- exp(m[i]) 
    m[i] <- bT[treatment[i]] + aP[population[i]]
    
  }

  # priors
  
  for (j in 1:Ntreatment) {
    
    k[j] ~ dgamma(0.01, 0.01) 
    bT[j] ~ dnorm(3, 0.5) 

  }
  
  for(r in 1:Npopulation) {
    
    aP[r] ~ dnorm(0, tauP) # population effect
    
  }
   
  tauP <- 1/(sigP * sigP) 
  sigP ~ dunif(0, 100)
  
}
```

## Run the NB model ##

```{r}
NB_run <- jags.fit(
  data = burden_jags,
  params = c("bT", "aP", "k"),
  model = NB_model,
  n.chains = 4)
```

## NB predictions versus experimental observations of endemic prevalence ##

Use the fitted model `NB_run` to get distributions of mean infection burden and mean endemic prevalence for each replicate population in constant and variable temperature treatments. These distributions will be derived by "simulating" the experiment (i.e., 26 time steps, 6 replicates per treatment). 

```{r}
(nb_posteriors <- 
  as.data.frame(MCMCchains(NB_run)) %>% 
  mutate(
    chain = c(rep(1, 5000), 
              rep(2, 5000), 
              rep(3, 5000), 
              rep(4, 5000)),
    observation = rep(1:5000, 4)
    ) %>% 
   mutate(across(1:14, exp)) %>% 
  gather(key = "parameter", value = "value", 1:16))

replicate_population <- as.data.frame(MCMCchains(NB_run)) %>% 
  select(1:12) %>% 
  names()

treatment <- as.data.frame(MCMCchains(NB_run)) %>% 
  select(13:14) %>% 
  names()

overdispersion <- as.data.frame(MCMCchains(NB_run)) %>% 
  select(15:16) %>% 
  names()
```

Initialize storage objects. 
```{r}
prevalence_nbpreds <- matrix(NA, nrow = 26, ncol = 12)
burden_nbpreds <- matrix(NA, nrow = 26, ncol = 12)
```

Calculate the number of individuals to sample from each replicate population in each treatment at each time step.

```{r}
nt_samples %>% 
  glimpse()
``` 

Sample from the fitted NB model using the following steps coded in the for loop:

1. Randomly choose a chain in the constant or variable treatment.
2. Sample a random observation in the chosen chain to get a set of parameters (m, k) for a given replicate population. 
4. Use the set of parameters to draw a sample of individuals from the fitted ZINB distribution. 
5. From the sample calculate endemic prevalence and mean infection burden. 

```{r, eval = FALSE}
for (i in 1:length(replicate_population)) {
    for (m in 1:26) {

      # for rep population i: 
      # choose set of estimated parameters from ZINB model fit
      
      if (i <=6) { # constant populations
      nb_params <- nb_posteriors %>%
        filter(chain == sample(1:4, 1, replace = FALSE)) %>%
        filter(observation == sample(1:5000, 1, replace = FALSE)) %>%
        filter(
          parameter %in% c(replicate_population[i],
                           treatment[1],
                           overdispersion[1]
                           )) %>%
        select(value) %>%
        unlist() 
      } 
    
      else { # variable populations
          
           nb_params <- nb_posteriors %>%
        filter(chain == sample(1:4, 1, replace = FALSE)) %>%
        filter(observation == sample(1:5000, 1, replace = FALSE)) %>%
        filter(
          parameter %in% c(replicate_population[i],
                           treatment[2],
                           overdispersion[2]
                           )) %>%
        select(value) %>%
        unlist() 
           
           }
      
      # sample from the fitted ZINB distribution
      samples <- rnbinom(
        nt_samples[m, i],
        size = nb_params[3],
        mu = nb_params[2] + nb_params[1]
      )
      
      # compute endemic prevalence and mean infection burden
      prevalence_nbpreds[m, i] <- (samples[samples != 0] %>% 
                                  length()) / nt_samples[m, i]
      
      burden_nbpreds[m, i] <- samples %>% mean()
      
}}
```

## Save the NB model for figures S ##

## Save the ZINB model for figures 4 and 5 ##

```{r eval = FALSE}
save(ZINB_run, # zinb model fit 
     zinb_posteriors, zinb_prevalence, zinb_burden, zinb_fitted, # treatment-level
     prevalence_zinbpreds, burden_zinbpreds, # replicate-level,
      prevalence_nbpreds, burden_nbpreds, # replicate-level
     file = "/Users/leilakrichel/Chp1/RData_Manuscript/ZINBExp_Fig4.RData")
```

# END #


